Park run with exponential Bradley-Terry

### 2025-04-09 inst/parkrun_idea.Rmd
page reference
another reference to underdispersion
more work on issue #239
consistentification

---
title: "Underdispersion in the context of community athletics"
output: bookdown::html_document2
author: "Robin K. S. Hankin"
bibliography: "hyper2.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017_rmd.

```{r deff}
library("hyper2")

f <- function(n,alpha){
  jj <- 1/alpha^(1:n)
  return(jj/sum(jj))
}

getlike <- function(v){
  alpha <- v[1]
  beta <- v[2]
 c(
tabulate(replicate(1000,which(rrace(setNames(f(26,alpha),letters))==letters[beta])),nbins=26)
,alpha,beta)
}
```

```{r calculateLikemat,cache=FALSE}
alpha_try <- seq(from=1,to=6,len=20)
beta_try <- 1:26
calculate_from_scratch <- FALSE
if(calculate_from_scratch){
    M <- as.matrix(expand.grid(alpha=alpha_try, beta=beta_try))
    system.time(L <- t(apply(M,1,getlike)))
    colnames(L) <- c(letters,'alpha','beta')
    write.table(L,file="likelihoodmatrix.txt")
} else {
    L <- read.table("likelihoodmatrix.txt")
}
L_integer <- L  # counts
L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
head(L_integer)
```

```{r makecontL}
colSums(L)
mL <- function(i){matrix(L[,i],length(alpha_try),length(beta_try))}
filled.contour(alpha_try,beta_try,mL(1))
filled.contour(alpha_try,beta_try,mL(5))
filled.contour(alpha_try,beta_try,mL(10))
filled.contour(alpha_try,beta_try,mL(15))
filled.contour(alpha_try,beta_try,mL(20))
filled.contour(alpha_try,beta_try,mL(26))
```

```{r}
#plot(table(L_integer[,1:26]))
```

## Now 10 in-silico races:
```{r doinsilico,cache=FALSE}
alphatrue <- 2.3
betatrue <- 4  
set.seed(1)
D <- replicate(10,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
sort(D)
```

```{r trylikeinsilico,cache=FALSE}
likelihood <- apply(L[,D],1,prod) + 0.00
likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
```

```{r showlike}
round(alpha_try,2)
matplot(alpha_try,likelihood,type='b')
matplot(beta_try,t(likelihood),type='b')
matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
filled.contour(alpha_try,beta_try,(likelihood))
table(likelihood==0)
plot(f(26,alphatrue))
abline(v=betatrue)
contour(alpha_try,beta_try,likelihood)
abline(v=alphatrue)
abline(h=betatrue)
contour(alpha_try,beta_try,log(likelihood) - log(max(likelihood)),levels=-(0:10))
abline(v=alphatrue)
abline(h=betatrue)
```

\subsection{Park run}

"Parkrun" is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 15 parkruns to date, and
from the first pair of numbers we see that 238 runners attended that
particular parkrun, of whom the author placed 173.

\begin{table*}[t]
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{ccccccccccc}
\hline
rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
\hline
\end{tabular}
\end{table*}

We may consider the author to have an unknown generalized Bradley
Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
two units of support~\cite{edwards1992} shown as a dotted line
illustrating a support interval of about 0.315 - 0.567; we may be
reasonably confident that the author's true BT strength lies in this
range.  In particular, note that a reasonable $H_0\colon
a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
susceptible to very minor improvement [$\simeq 0.351$] by moving to
the evaluate.

```{asis}
\begin{figure}[t]
\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
\caption{Author's strength\label{parkrunsupport}}
\end{figure}
```

>>>>

---
title: "Park run with exponential Bradley-Terry"
author: "robin hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
set.seed(0)
```

Basic idea is to have $n$ competitors with BT strengths $\beta, \beta
x,\ldots \beta x^{n-1}$ with $x\in[0,1]$.  The unit sum constraint
gives $\beta=\frac{1-x}{1-x^n}$ but this usually cancels out.  One of
the competitors is the _focal_ competitor with BT strength $\beta
x^{r-1}$ for some $r$ with $1\leqslant r < n$.  Thus $r=1$ means that
the focal competitor is the strongest among the $n$, $r=2$ means he is
the second, and so on.

```{r defineBT}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- paste0("p",str_pad(1:n,ceiling(log10(n))))
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:

```{r showtabulate}
(jj <- BT(9,0.5))
o <- tabulate(replicate(1e4,which(rrace(jj) == "p4")))
o
plot(o,type='h')
```

```{r definegetprob, cache=TRUE}
getprob <- function(n, r, x, N=1e4){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
```

We translate $\alpha\in [0,1]$ to a number from 1 to $n$ inclusive
with $\lceil n\alpha\rceil$

Write a likelihood function for 9,12 [that is, came 9th in a class of
12] There are 8 better and 3 worse than the focal student.

```{r ninetwelve}
single_like_fun <- function(n, a,          # n competitors, places a [e.g a=3 -> third]
                            a_try, x_try,  # values to try
                            N=1e2){        # N is number of trials
    p_try <- ceiling(n * a_try)
    jj <- as.matrix(expand.grid(p_try, x_try))
    II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N)}))
    matrix(II[, ceiling(a*n) + 2], length(a_try), length(x_try))
}
```

```{r makesupp912, cache=TRUE}
a_try <- seq(from = 0.01, to = 0.99, len = 10)
x_try <- seq(from = 0.2, to = 1, by = 0.1)
```

In the following chunks, each execution of
`single_like_fun(...,N=1000)` takes  about 5 minutes

```{r sethowmany}
howmany <- 100
```

```{r parkrunlikefunobs1, cache = TRUE}
print(system.time(parkrun01 <- single_like_fun(259, 177/259, N = howmany, a_try, x_try)))
```

```{r}
parkrun01
```

```{r parkrunlikefunobs2, cache = TRUE}
 print(system.time(parkrun02 <- single_like_fun(305, 222/305, N = howmany, a_try, x_try)))
```

```{r}
parkrun02
```

```{r parkrunlikefunobs3, cache = TRUE}
print(system.time(parkrun03 <- single_like_fun(297, 206/297, N = howmany, a_try, x_try)))
```

```{r}
parkrun03
```

```{r parkrunlikefunobs4, cache = TRUE}
print(system.time(parkrun04 <- single_like_fun(241, 142/241, N = howmany*10, a_try, x_try)))
```

```{r}
parkrun04
```

```{r parkrunlikefunobs5, cache = TRUE}
print(system.time(parkrun05 <- single_like_fun(179, 118/179, N = howmany, a_try, x_try)))
```

```{r}
parkrun05
```

```{r parkrunlikefunobs6, cache = TRUE}
print(system.time(parkrun06 <- single_like_fun(338, 224/338, N = howmany, a_try, x_try)))
```

```{r}
parkrun06
```

```{r parkrunlikefunobs7, cache = TRUE}
print(system.time(parkrun07 <- single_like_fun(203, 128/203, N = howmany, a_try, x_try)))
```

```{r}
parkrun07
```

```{r}
c(
sum(parkrun01),
sum(parkrun02),
sum(parkrun03),
sum(parkrun04),
sum(parkrun05),
sum(parkrun06),
sum(parkrun07)
)
```

```{r}
contour(a_try,x_try,parkrun05)
contour(a_try,x_try,parkrun06)
parkrun01*parkrun02*parkrun03*parkrun04*parkrun05*parkrun06*parkrun07
```
